Recursive partitioning of linear growth curve models

### 2022-12-10 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Updated LGCM GLMM-tree paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3241 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ad9f2ca5e6d0d5ed9e683530438301d2e98e11c0

\documentclass[doc,floatsintext,natbib]{apa7}
%\usepackage[man,floatsintext]{apa6}
%\usepackage[doc,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
%\usepackage[colorinlistoftodos]{todonotes}
%%\todo[inline, color=green!40]{This is an inline comment.}

@@

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Between-subject heterogeneity is commonly encountered and often of specific research interest. We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models. GLMM trees were originally proposed for clustered cross-sectional data; here we introduces extensions for longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate the performance of GLMM trees in partitioning growth curve models on simulated datasets and existing datasets on children's development of reading, math and science abilities. We compare the performance of GLMM trees with the SEM trees, LongCART and longRPart methods. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for analysing discrete as well as continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

Modeling development over time is of prime interest in psychological research. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter-individual heterogeneity is likely encountered in such studies. For example, some (groups of) participants may show higher or lower rates of growth over time. Development over time is generally studied with mixed-effects or latent-variable models, which allow for capturing inter-individual heterogeneity and explaining this heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.

However, the covariates that predict or affect developmental trajectories may not be known in advance and often researchers may have a specific interest in identifying these covariates. Recursive partitioning methods (RPMs) are exploratory methods that allow for identifying relevant predictors from a potentially (very) large number of covariates. At the same time, RPMs allow for direct identification of heterogeneous subgroups. Several existing RPMs allow for subgroup detection in growth-curve models: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Further, longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models \footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

In this paper, we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for subgroup detection in LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. Yet, GLMM trees provide a much more general method, allowing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data often come with two central characteristics that may require a different estimation approach: Higher random-effects variance (i.e., higher intra-class correlations) and predictors of interest measured at higher levels (e.g., time-constant covariates). In this paper, we introduce and test two recent extensions of GLMM trees that can account for these characteristics. We focus on the specific use case of partitioning LGCMs, but methods for dealing with stronger intra-class correlations and higher-level covariates of interest are obviously of relevance beyond the realm of LGCMs.

\subsection{Recursive partitioning methods for longitudinal data}

Compared to the aforementioned RPMs for partitioning LGCMs \citep{Loh02, AbdoyLeBl02, Lee05, SuyMene11, BranyOert13, ArnoyVoel21, EoyCho14, KundyHare19}, GLMM trees differ in that they do not fit a full parametric model in each of the terminal nodes. GLMM trees estimate the fixed-effects parameters locally, using only the observations within a terminal node, and estimate the random-effects parameters globally, using all observations. Not only does this reduce computation time, it also reduces model complexity, which in turn may reduce instability and overfitting. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes. In effect, the approach replaces the fixed-effects part of the mixed-effects model with a decision tree with constant fits in the terminal nodes, while the random effects are estimated as usual. \cite{FokkySmit18} generalized this approach to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of interaction effects.

The computational advantage of GLMM trees is highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These latter methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, a likelihood ratio test is computed for every possible cut point. In addition to a heavy computational burden, this also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces the computational burden, and allows for separating variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

\subsubsection{Current paper}

In what follows, the next section explains how GLMM trees are estimated, and how it can be adapted to better account for longitudinal data structures. In three empirical evaluations, we assess and compare performance of GLMM trees in partitioning LGCMs: In Simulation Study I, we assess performance of the proposed adaptations to GLMM trees. In Simulation Study II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in partitioning existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and their implications.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. In principle, generalizations to other distributions within the GLM. Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises two columns: a column of 1s for the intercept and a column of observed timepoints for subject $i$. The number ($N_i$) and values of observed timepoints may differ between subjects. While in a traditional GLMM, the fixed-effects parameters are assumed constant over all observations, but in the GLMM tree model, $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$ and $b_i$ is a vector of random effects for subject $i$. Finally, we assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

@@

Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed found only minor impact of different initialization values on the final model, which became smaller with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, repeated measures on the same subjects tend to be correlated more strongly than observations nested withiin the same unit in cross-sectional data. If the random effects are in fact sizable, assuming them to be zero could provide an unrealistic starting point that is difficult to overcome in subsequent iterations. Our first hypothesis is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure improves accuracy of GLMM trees.

@@

\begin{eqnarray}
\label{eq:efp}
W_{k}(t) = \hat{J}^{-1/2} n^{-1/2} \sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
\end{eqnarray}

where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n$ gives the number of observations in the current node. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.

\textcolor{red}{Perhaps replace $n$ with $N_j$? This is more line with notation of GLMM tree model, but might be confusing with cov mat $J$?}

@@

When partitioning longitudinal data, covariates will often be measured at the subject level. This clustering of observations within subjects can be accounted for in computing a so-called \textit{clustered} covariance matrix $\hat{J}$ \citep[e.g., ][]{ZeilyKoll20}. This resembles a GEE-type approach with an independence correlation structure. Our second hypothesis is that for partitioning LGCMs, use of clustered covariances improves accuracy of GLMM trees.

@@

\subsubsection{Data-generating design}

<<echo=FALSE>>=
library("partykit")
set.seed(42)
U1 <- sample(0:1, size = 1250, replace = TRUE) 
U2 <- round(rnorm(1250))
U3 <- round(rnorm(1250))
time <- rep(0:4, each = 250)
## left-most node: x1 == 0 & x2 <= 0
## intercept = -1, slope = -1
node3 <- as.numeric(U1 == 0 & U2 <= 0)
## left-middle node: x1 == 0 & x2 > 0
## intercept = -1, slope = 0
node4 <- as.numeric(U1 == 0 & U2 > 0)
## right-middle node: x1 == 1 & x3 =< 0
## intercept = 1, slope = 0
node6 <- as.numeric(U1 == 1 & U3 <= 0)
## right-most node: x1 == 1 & x3 > 0
## intercept = 1, slope = 1
node7 <- as.numeric(U1 == 1 & U3 > 0)
y <- (node3 * -1 - node3 * time) + (-node4) +
 (node6) + (node7 * 1 + node7 * time) 
data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)

fig4 <- party(
  partynode(1L,
            split = partysplit(1L, breaks = 1),
            kids = list(
              partynode(2L,
                        split = partysplit(2L, breaks = 0),
                        kids = list(partynode(3L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '-1.0'))),
                          partynode(4L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))))),
              partynode(5L,
                        split = partysplit(3L, breaks = 0),
                        kids = list(
                          partynode(6L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))),
                          partynode(7L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '1.0'))))))),
  data.frame(data)
)
#plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)

@@

\begin{figure}[!t]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
<<fig=T, height=5, width=5, echo=F>>=
#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
#     tp_args = list(pointcol = "white", xlab = "time",
#                    mainlab = function(id, nobs) paste("node", id)))
plot(tree, tnex = 3.5, terminal_panel = combine_panels(
    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
                        mainlab = function(id, nobs) paste("node", id)),
    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
    party2 = fig4), 
  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@
\label{fig:design_tree}
\end{figure}

@@

\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part. The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively). The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints. The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$ \footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees.}. Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next. Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.

@@

\begin{itemize}
\setlength\itemsep{0em}
\setlength{\itemindent}{0.2in}
\item Number of subjects: small ($N = 100$) or large ($N = 250$).
\item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 2$).
\item Variance of the random slope: small ($\sigma_{b_1}^2 = .1$) or large ($\sigma_{b_1}^2 = .4$).
\item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
\item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = .3$).
\end{itemize}

We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We used R \citep[version 4.1.2; ][]{R22} to generate and analyze all data.

\subsubsection{Partitioning approaches}

We fitted ten trees to each dataset:

\begin{itemize}
  \item Two linear model (LM) trees: 
  \begin{itemize}
    \item A tree employing default settings, where the parameter stability tests use observation-level covariances.
    \item A tree where the parameter stability tests use clustered covariances. 
  \end{itemize}
  \item Eight linear mixed model (LMM) trees: 
  \begin{itemize}
    \item Two trees fitted using default settings, where parameter stability tests use observation-level covariances and estimation is initialized with the tree structure. 
    \item Two trees where the parameter stability tests use cluster-level covariances.
    \item Two trees where estimation is initialized with the random effects. 
    \item Two trees where the parameter stability tests use cluster-level covariances and estimation is initialized with the random effects.
  \end{itemize}
\end{itemize}

Each of the two LMM trees were fitted with different specifications of the random effects: One of the LMM trees estimated random intercepts only, one of the LMM trees estimated both random intercepts and slopes. Because variances of random intercepts and slopes were $> 0$ in all cells of the design, the three different random-effects specifications (LM trees assume 0 random effects) allow for assessing the effect of mis-specifying the random effects. To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}. To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}.

@@

We evaluated tree accuracy by counting the number of splits in each tree. Trees with $> 3$ splits are indicative of Type-I error, while trees with $> 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables). We also assessed which variable was selected for the first split in every tree.

@@

M_default <- mean(tmp$tree_size[tmp$default_ids])
M_cluster <- mean(tmp$tree_size[tmp$cluster_ids])
M_ranefstart <- mean(tmp$tree_size[tmp$ranefstart_ids])
M_both <- mean(tmp$tree_size[tmp$both_ids])
@

@@

<<echo=FALSE>>=
means <- tapply(tmp$tree_size, tmp$method_id, mean)
@

\begin{figure}[!ht]
\caption{Tree size distributions for LM(M) trees.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=3.5>>=
## Prepare data
tmp <-  tree_sizes_long
tmp$tree_size <- (tmp$tree_size-1)/2
tmp <- tmp[-which(tmp$method_id == "true") , ]
tmp$method_id <- factor(tmp$method_id)

>>>>

\documentclass[doc,floatsintext,natbib]{apa7}
%\usepackage[man,floatsintext]{apa6}
%\usepackage[doc,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{setspace}

\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references

%\usepackage[colorinlistoftodos]{todonotes}
%%\todo[inline, color=green!40]{This is an inline comment.}

@@

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is likely encountered in such models and researchers are often interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models. GLMM trees were originally proposed for clustered cross-sectional data; here we extend the method to allow for longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities. We compare the performance of GLMM trees with three alternative methods: SEM trees, LongCART and longRPart. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

Development over time is of prime interest in psychological researchers. For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}. Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}. However, the covariates that are predictive of development may often not be known in advance, and identifying them may be of prime interest.

Recursive partitioning methods (RPMs) allow for identifying relevant predictors from a potentially (very) large number of covariates. Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

Here we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for partitioning LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. Yet, GLMM trees provide a more general method 2for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two recent extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wide range of settings where covariates are measured at higher levels and/or random-effects variance is substantial.

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes. Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \cite{FokkySmit18}.

All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based the observations in that node only. This fully local estimation approach brings higher computational load and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, a likelihood ratio test is computed for every possible cut point. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection.

In summary, compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.

In what follows, we first describe estimation of GLMM trees, and possible adjustments to better account for longitudinal data structures. Next, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees. In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and implications.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Thus, in the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

@@

Studies to date indicate that initializing estimation with zero random effects yields accurate final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found minor impact that became smaller with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be stronger: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that is difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy.

@@

\begin{eqnarray}
\label{eq:efp}
W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
\end{eqnarray}

where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n_j$ gives the number of observations in the current node. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the estimated scores, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.

@@

When partitioning longitudinal data, covariates will often be measured at the subject level, which should be accounted for in computing the estimated covariance matrix $\hat{J}$. In general, this computation makes use of the scores. By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}. This resembles a GEE-type approach with an independence correlation structure. Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve accuracy of the tree.

% $\textrm{Var} \hat{\beta}$ = (X^\top X)^{-1} X^\top \Omega X (X^\top X)^{-1}$

% The $X^\top \Omega X$ is the meat, and the $(X^\top X)^{-1}$ is the bread. For computing robust standard errors, we compute a new kind of meat. $\Omega = \sigma^2 I_n$, a matrix with $\sigma^2$ on the diagonal and zeros elsewhere. $n$ can be the number of observations, or the number of clusters.

@@

\subsubsection{Data generation}

<<echo=FALSE>>=
library("partykit")
set.seed(42)
U1 <- sample(0:1, size = 1250, replace = TRUE) 
U2 <- round(rnorm(1250))
U3 <- round(rnorm(1250))
time <- rep(0:4, each = 250)
## left-most node: x1 == 0 & x2 <= 0
## intercept = -1, slope = -1
node3 <- as.numeric(U1 == 0 & U2 <= 0)
## left-middle node: x1 == 0 & x2 > 0
## intercept = -1, slope = 0
node4 <- as.numeric(U1 == 0 & U2 > 0)
## right-middle node: x1 == 1 & x3 =< 0
## intercept = 1, slope = 0
node6 <- as.numeric(U1 == 1 & U3 <= 0)
## right-most node: x1 == 1 & x3 > 0
## intercept = 1, slope = 1
node7 <- as.numeric(U1 == 1 & U3 > 0)
y <- -(node3 + node3*time) - node4 + node6 + (node7 + node7*time) 
data <- data.frame(U1 = as.factor(U1), U2, U3, time, y)   
tree <- lmtree(y ~ time | U1 + U2 + U3, data = data, maxdepth = 3)
fig <- party(
  partynode(1L,
            split = partysplit(1L, breaks = 1),
            kids = list(
              partynode(2L,
                        split = partysplit(2L, breaks = 0),
                        kids = list(partynode(3L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '-1.0'))),
                          partynode(4L, info = c(
                            expression(beta[j0] == '-1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))))),
              partynode(5L,
                        split = partysplit(3L, breaks = 0),
                        kids = list(
                          partynode(6L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '0.0'))),
                          partynode(7L, info = c(
                            expression(beta[j0] == '1.0'),
                            expression(''),
                            expression(beta[j1] == '1.0'))))))),
  data.frame(data)
)

@@

\begin{figure}[bt]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
<<fig=T, height=5, width=5, echo=F>>=
plot(tree, tnex = 3.5, terminal_panel = combine_panels(
    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
                        mainlab = function(id, nobs) paste("node", id)),
    panel2 = node_terminal(fig, FUN = identity, height = 4, width = 12, id = FALSE),
    party2 = fig), 
  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@
\label{fig:design_tree}
\end{figure}

@@

\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part. The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively). The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints. The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$\footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees in the next section.}. Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next. Values of $\epsilon_{i}$ were independently generated from a normal distribution with a mean of zero and a variance of 5.

@@

\begin{itemize}
\setlength\itemsep{0em}
\setlength{\itemindent}{0.2in}
\item Number of subjects: small ($N = 100$) or large ($N = 250$).
\item Variance of the random intercept: small ($\sigma_{b_0}^2 = 1$) or large ($\sigma_{b_0}^2 = 2$).
\item Variance of the random slope: small ($\sigma_{b_1}^2 = 0.1$) or large ($\sigma_{b_1}^2 = 0.4$).
\item Number of noise partitioning variables: small ($p = 5$) or large ($p = 25$).
\item Intercorrelation between partitioning variables: absent ($\rho = 0$) or present ($\rho = .3$).
\end{itemize}

We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We performed all data generation and analysis in R \citep[version 4.1.2; ][]{R22}.

\subsubsection{Model fitting}

We applied ten partitioning approaches to every generated dataset, each employing one of three different random-effects specifications:

    \item $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. That is, random effects and thus their variance were fixed to 0, and two fixed-effects linear model (LM) trees were fitted.
    \begin{itemize}
        \item The default observation-level covariances were used to compute the parameter stability tests.
        \item Clustered covariances were used to compute the parameter stability tests. 
    \end{itemize}

    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$. That is, four linear mixed-effects model (LMM) tree were fitted in which the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0: 
    \begin{itemize}
      \item Default settings were used, in which parameter stability tests employed observation-level covariances and estimation is initialized with the tree structure. 
    \item Instead, clustered covariances were used to compute the parameter stability tests.
    \item Instead, estimation was initialized with the random effects. 
    \item Instead, clustered covariances were used to compute the parameter stability tests and estimation was initialized with the random effects.
  \end{itemize}

    \item $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$. That is, four LMM trees were fitted in which the variance of both the random intercept and slope were freely estimated. The same four approaches were applied (default, clustered covariances, random-effects initialization, clustered covariances and random-effects initialization) as with the previous LMM tree specification.
  \end{itemize}

To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}. To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}. To compute clustered covariance, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}. We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.

@@

We evaluated tree accuracy by counting the number of splits in each tree. Trees with $> 3$ splits are indicative of Type-I error, while trees with $> 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables). We also assessed the variable selected for the first split in every tree.

@@

M_default <- mean(tmp$tree_size[tmp$default_ids])
M_cluster <- mean(tmp$tree_size[tmp$cluster_ids])
M_ranefstart <- mean(tmp$tree_size[tmp$ranefstart_ids])
M_both <- mean(tmp$tree_size[tmp$both_ids])

means <- tapply(tmp$tree_size, tmp$method_id, mean)
sds <- tapply(tmp$tree_size, tmp$method_id, sd)
mads <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(abs(x-3)))
props_more_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x > 3L))
props_less_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x < 3L))
props_corr_splits <- tapply(tmp$tree_size, tmp$method_id, function(x) mean(x == 3L))
@

@@

\begin{figure}[!ht]
\caption{Tree size distributions for LM trees (left panel) and LMM trees (middle and right panel).}
\begin{subfigure}{1.25\textwidth}
<<echo=FALSE, fig=TRUE, height=4>>=
library("dplyr")

## Prepare data
tmp <-  tree_sizes_long
tmp$tree_size <- (tmp$tree_size-1)/2 ## count splits instead of nodes
tmp <- tmp[-which(tmp$method_id == "true") , ]
tmp$method_id <- factor(tmp$method_id)
